---
title: "Build and deploy a stroke prediction model using R"
date: "`r Sys.Date()`"
output: html_document
author: "Muhammad Jawwad Kiani!"
---

# About Data Analysis Report

This RMarkdown file contains the report of the data analysis done for the project on building and deploying a stroke prediction model in R. It contains analysis such as data exploration, summary statistics and building the prediction models. The final report was completed on `r date()`.

**Data Description:**

According to the World Health Organization (WHO) stroke is the 2nd leading cause of death globally, responsible for approximately 11% of total deaths.

This data set is used to predict whether a patient is likely to get stroke based on the input parameters like gender, age, various diseases, and smoking status. Each row in the data provides relevant information about the patient.

# Task One: Import data and data preprocessing

## Load data and install packages

```{r}
# Install necessary packages if not already installed
list.of.packages <- c("tidyverse", "caret", "randomForest", "e1071", "ROCR")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

# Load libraries
library(tidyverse)
library(caret)
library(randomForest)
library(e1071)
library(ROCR)

# Load the dataset
healthcare_dataset_stroke_data <- read.csv("healthcare-dataset-stroke-data.csv", stringsAsFactors = FALSE)


stroke_data <-healthcare_dataset_stroke_data

# View the first few rows
head(stroke_data)
```

## Describe and explore the data

```{r}
# Summary statistics
summary(stroke_data)

# Check for missing values
colSums(is.na(stroke_data))

# Visualize distribution of key variables
ggplot(stroke_data, aes(x=age)) + geom_histogram(binwidth=5, fill="blue", color="black") + theme_minimal()

# Stroke vs Age boxplot
ggplot(stroke_data, aes(x=factor(stroke), y=age, fill=factor(stroke))) +
  geom_boxplot() + labs(x="Stroke (0 = No, 1 = Yes)", y="Age") + theme_minimal()

# Gender distribution
ggplot(stroke_data, aes(x=gender, fill=factor(stroke))) +
  geom_bar(position="fill") + labs(y="Proportion", fill="Stroke") + theme_minimal()

```
```{r}
# Step 1: Check for non-numeric BMI values using regex
invalid_bmi <- unique(stroke_data$bmi[!grepl("^[0-9.]+$", stroke_data$bmi)])

# Step 2: Display a clean message depending on findings
if (length(invalid_bmi) == 0) {
  print("No invalid BMI values found.")
} else {
  print(paste("Invalid BMI values:", paste(invalid_bmi, collapse = ", ")))
}

# Step 3: Replace all invalid BMI values with NA
stroke_data$bmi[!grepl("^[0-9.]+$", stroke_data$bmi)] <- NA

# Step 4: Convert BMI column to numeric
stroke_data$bmi <- as.numeric(stroke_data$bmi)

# Step 5: Impute missing BMI values with the median
stroke_data$bmi[is.na(stroke_data$bmi)] <- median(stroke_data$bmi, na.rm = TRUE)

# Step 6: Confirm no NA values remain in any column
print("Missing values per column:")
print(colSums(is.na(stroke_data)))

# Step 7: Optional check – confirm bmi is numeric and valid
print(str(stroke_data$bmi))
print(summary(stroke_data$bmi))








```
# Task Two: Build prediction models

```{r}
# Set seed for reproducibility
set.seed(123)

# Ensure stroke is a factor (for classification)
stroke_data$stroke <- as.factor(stroke_data$stroke)

# Split data into train (70%) and test (30%) sets
trainIndex <- createDataPartition(stroke_data$stroke, p = 0.7, list = FALSE)
train_data <- stroke_data[trainIndex, ]
test_data <- stroke_data[-trainIndex, ]

# Logistic Regression Model
log_model <- glm(stroke ~ ., data = train_data, family = binomial)

# Random Forest Model (Classification)
rf_model <- randomForest(stroke ~ ., data = train_data, ntree = 100, importance = TRUE)

# Support Vector Machine (SVM) Model
svm_model <- svm(stroke ~ ., data = train_data, probability = TRUE)

# Confirm models are trained
print("Models trained successfully.")

```

# Task Three: Evaluate and select prediction models

```{r}
# Logistic Regression Predictions and ROC
log_pred_prob <- predict(log_model, test_data, type="response")
log_pred <- ifelse(log_pred_prob > 0.5, 1, 0)

# Random Forest Predictions
rf_pred <- predict(rf_model, test_data)
rf_pred_prob <- predict(rf_model, test_data, type="prob")[,2]

# SVM Predictions
svm_pred <- predict(svm_model, test_data)
svm_pred_prob <- attr(predict(svm_model, test_data, probability=TRUE), "probabilities")[,2]

# Confusion Matrix for Logistic Regression
confusionMatrix(as.factor(log_pred), test_data$stroke, positive="1")

# Confusion Matrix for Random Forest
confusionMatrix(rf_pred, test_data$stroke, positive="1")

# Confusion Matrix for SVM
confusionMatrix(svm_pred, test_data$stroke, positive="1")

# ROC Curve for all models
pred_log <- prediction(log_pred_prob, test_data$stroke)
perf_log <- performance(pred_log, "tpr", "fpr")

pred_rf <- prediction(rf_pred_prob, test_data$stroke)
perf_rf <- performance(pred_rf, "tpr", "fpr")

pred_svm <- prediction(svm_pred_prob, test_data$stroke)
perf_svm <- performance(pred_svm, "tpr", "fpr")

plot(perf_log, col="blue", main="ROC Curves for Stroke Prediction Models")
plot(perf_rf, col="green", add=TRUE)
plot(perf_svm, col="red", add=TRUE)
legend("bottomright", legend=c("Logistic Regression","Random Forest","SVM"),
       col=c("blue","green","red"), lty=1)

```

# Task Four: Deploy the prediction model

```{r}

saveRDS(rf_model, file = "stroke_rf_model.rds")


```

# Task Five: Findings and Conclusions

Data Overview:

Dataset contains 5110 observations with features including age, gender, hypertension, heart disease, marital status, work type, residence type, average glucose level, BMI, smoking status, and stroke occurrence.

The target variable stroke is highly imbalanced, with about 4.8% of cases labeled as stroke (stroke = 1).

Data Preprocessing:

Categorical variables were correctly converted to factors.

BMI values had some invalid or missing entries which were identified and imputed with the median BMI to maintain data integrity.

Missing values were handled, and no NA values remained after cleaning.

Modeling and Performance:

Logistic Regression, Random Forest, and SVM models were trained on a 70-30 train-test split.

The Random Forest model was saved successfully on your local machine.

From the confusion matrices and classification reports:

The model shows very high accuracy (~95%) but this is misleading because of class imbalance.

Sensitivity (Recall for stroke cases) was extremely low (around 0% to 1.3%), meaning the model rarely correctly predicts actual stroke cases.

Specificity (true negative rate) was 100%, meaning non-stroke cases were almost always correctly identified.

This results in a poor ability to detect strokes despite high overall accuracy, a common issue with imbalanced data.

Kappa statistics near zero indicate the model predictions are barely better than random guessing on the minority class.

ROC Curve:

The ROC curve and AUC (not explicitly shown here but typically evaluated) would help assess discriminatory power better than accuracy alone.

Given low sensitivity, the AUC likely is modest, reflecting the challenge of predicting the minority class.

Conclusions:
Model performance is heavily influenced by class imbalance. Accuracy alone is not a good metric here because the dataset is dominated by non-stroke cases.

The current models tend to predict the majority class (no stroke) almost exclusively, missing most stroke cases.

To improve stroke prediction, consider:

Applying techniques to handle imbalance, such as SMOTE (Synthetic Minority Over-sampling Technique), class weighting, or balanced random forests.

Using evaluation metrics focused on the minority class, such as Precision, Recall, F1-score, or AUC-PR (Area Under Precision-Recall Curve).

Exploring additional predictive features or domain-specific feature engineering.

Possibly using ensemble methods or tuning hyperparameters to boost minority class detection.

Data cleaning and preprocessing, especially handling invalid BMI values, was successfully performed to ensure model input quality.


